### 2023-11-29 thesis/MSECE_thesis.tex
Nearly final version of thesis.

\usepackage{cite}
\usepackage{listings}
\usepackage{multicol}
\usepackage{newtxtext}
\usepackage{setspace}
\usepackage{sectsty}
\usepackage{booktabs}
\usepackage[T1]{fontenc}
\usepackage[numbib]{tocbibind}
\usepackage[lmargin=1.25in,rmargin=1.25in,tmargin=1.0in,bmargin=1.0in]{geometry}

@@

    \normalsize
    by\\
    \textit{Aaron P.~Van De Brook}\\

@@

% Signature lines/committee members
\begin{center}
    \begin{minipage}{3in}
        \vspace*{0.25in}
        THESIS COMMITTEE:\\
        \vspace*{0.25in}
        \hrule
        \vspace*{2pt}
        Jianhua Liu, Ph.D.\\
        Committee Chairman\\
        \vspace*{0.75in}
        \hrule
        \vspace*{2pt}
        Prashant Shekhar, Ph.D.\\
        Committee Member\\
        \vspace*{0.75in}
        \hrule
        \vspace*{2pt}
        Andrew Schneider, Ph.D.\\
        Committee Member\\
        \vspace*{0.75in}
        \hrule
        \vspace*{2pt}
        Farahzad Behi\\
        Department Chair, Electrical Engineering \& Computer Science\\
        \vspace*{0.75in}
        \hrule
        \vspace*{2pt}
        Associate Vice President for Academics
    \end{minipage}
\end{center}
\newpage
\tableofcontents
\newpage
\doublespacing{}

\section{Abstract}\label{sec:abstract}

\section{Introduction}\label{sec:introduction}
There have been noticeable and significant advancements in the general domain of artificial intelligence recently. Automatic speech recognition and
language modeling technologies, in particular, have made tremendous advances over the past two decades achieving performance levels good enough to
be brought to consumer facing products such as personal assistants, search engines, phones, televisions, etc. Due to the success in the general
domain, applications have begun to spin-off into problem-specific domains, such as (and most notably for this work) aviation. For example, several
speech recognition systems with rule-based natural language inference and semantic extraction algorithms have been implemented at EUROCONTROL
simulation centers across Europe.

At the time of this writing, the current focus of general machine learning algorithms in the automatic speech recognition (ASR) and language modeling
domains has been a mixture of deep learning methods, namely; recurrent, convolutional, and transformer neural networks due to the availability of
large corpora such as LibriSpeech~\cite{panayotov_librispeech_2015} and Mozilla Common Voice~\cite{ardila_common_2020} for ASR and
WikiText~\cite{merity_pointer_2016} and IMDB~\cite{maas_learning_2011} corpora for language modeling. In contrast to the general domain, the aviation
domain still prefers, and sees much success with traditional machine learning models for both ASR and language modeling applications. Typically, the
most prevalent of the traditional models are Hidden Markov models and Gaussian Mixture models, in conjunction with word lattices, N-grams, and/or
rule-based approaches depending on the application and operating environment of the models. The most likely reason for this divergence between the
domains is the difference and relative lack of labeled data in the aviation domain. By comparison, the largest corpus in the general domain is
LibriSpeech with approximately 1000 hours of labeled data whereas, for aviation, the largest corpus is Air Traffic Control Complete with approximately
26 hours of labeled data\footnote{Air Traffic Control Complete has approximately 70 hours of audio data, however, about 26 hours of the 70 are
    labeled.}.

@@

\section{Data Sources}\label{sec:data_source}
There are four datasets used in this work. The datasets are primarily speech corpora intended either for automatic speech recognition
research in the domain of aviation communications, linguistics research in aviation english, or both. As a result, the text-based data
samples that are used to train the language models in this work are transcripts of aural speech. The datasets used are described in more
detail below. All recordings and transcripts are in english, although there are multiple instances of foreign words occurring

\subsection{Air Traffic Control Complete}\label{sec:atcc}
The Air Traffic Control Complete (ATCC) corpus is a speech corpus consisting of audio recordings with corresponding transcriptions.
ATCC is a collection of three smaller corpora recorded at Dallas-Fort Worth, Logan International, and Washington National airports and
transcribed by subject matter experts (usually controllers and/or pilots) familiar with the terminology specific to their respective
areas\cite{godfrey_air_1994}.

\subsection{ATCO2}\label{sec:atco2}
The ATCO2 dataset is a speech corpus that consists of audio communications at Prague and Brno airports, in Czechia, along with corresponding
transcriptions. The speech recordings and transcripts are crowd sourced from volunteers\cite{szoke_detecting_2021}.

\subsection{Air Traffic Control Simulation}\label{sec:atcosim}
The Air Traffic Control Simulation (ATCOSIM) corpus is a speech corpus consisting of audio recordings and corresponding transcriptions.
The data was recorded at a air traffic control simulation center in Bretigny-sur-Orge, France, specifically the EUROCONTROL Experimental Centre.
In this data, only the controllers' voices are included and therefore the transcripts only include the controller's side of each interaction.
The audio data was transcribed by one person, trained according to the guidelines and formatting requirements of the corpus. After all data was
transcribed, it was reviewed and corrected and any remaining problems were reviewed by an operational air traffic controller and
resolved\cite{hofbauer_atcosim_2008}.

\subsection{ZCU CZ ATC Corpus}\label{sec:zcu_atc}
The ZCU CZ ATC corpus consists of audio recordings and corresponding transcripts in the Czech airspace. Both controller and pilot sides of the
communications are included in this data. Annotations were created by experienced transcribers/annotators and labeled samples were randomly selected
for review during the annotation process. After the dataset was completely labeled, all samples were checked, corrected, and
unified\cite{smidl_air_2019}.

@@

\begin{table}[h!]
    \centering
    \begin{tabular}{l p{0.65\linewidth}}
        \toprule
        Phenomenon               & Definition                                                                                                              \\
        \midrule
        Hesitations              & Instances of hesitation for which the specific realization is not defined (usually some special token is used instead). \\
        \midrule
        Partial \& cut-off words & Partially pronounced words are indicated by a hyphen character where the pronunciation of the word is missing.          \\
        \midrule
        Filled pauses            & Hesitations with a verbal component that has been described or otherwise defined by the corpus/annotator.               \\
        \bottomrule
    \end{tabular}
    \label{tab:phenomena_definitions}
    \caption{Speech/transcription phenomena with corresponding convention for transcribing or translating, depending on the corpus being processed.}
\end{table}

\begin{table}[h!]
    \centering
    \begin{tabular}{l l}
        \toprule
        Phenomenon                   & Convention for representation \\
        \midrule
        Hesitations                  & [HES]                         \\
        \midrule
        Partial \& cut-off words     & -                             \\
        \midrule
        Filled pauses (open-mouth)   & uh, um, ah                    \\
        \midrule
        Filled pauses (closed-mouth) & hm, mhm                       \\
        \bottomrule
    \end{tabular}
    \label{tab:phenomena_conventions}
    \caption{Conventions for transcript representation/translation of phenomena to keep transcripts consistent across corpora.}
\end{table}

@@

\section{Data Analysis}\label{sec:data_analysis}
% sequence lengths, unique tokens, total number of tokens, total number of samples, breakdown by each dataset
Relevant corpora statistics were observed or calculated and are presented in Table~\ref{table:corpora_stats}. This includes the number of samples in
the corpus, mean sequence length per sample in the corpus, the number of unique tokens in the corpus, and the total number of tokens in the corpus.
The same statistics were also calculated across corpora, since all four corpora will be combined for training and testing the language and speech
recognition models. Note that this is based on the raw data after loading and unifying the corpora, prior to preprocessing for model training/testing
and outlier removal.

\begin{table}[h!]
    \centering
    \begin{tabular}{l r r r r}
        \toprule
        \textbf{Corpus} & \textbf{Samples} & \textbf{Mean Sequence Length (\(\pm \sigma\))} & \textbf{Unique Tokens} & \textbf{Tokens} \\
        \midrule
        ATCC            & 29,862           & \(11.90 \pm 7.22\)                             & 2,209                  & 355,473         \\
        ATCO2           & 874              & \(12.28 \pm 4.12\)                             & 786                    & 10,733          \\
        ATCOSIM         & 9,544            & \(11.23 \pm 4.12\)                             & 823                    & 107,153         \\
        ZCU CZ ATC      & 14,435           & \(10.05 \pm 5.50\)                             & 2,983                  & 145,107         \\
        \midrule
        Total           & 54,715           & \(11.30 \pm 6.36\)                             & 5,040                  & 618,466         \\
        \bottomrule
    \end{tabular}
    \label{table:corpora_stats}
    \caption{The number of samples, mean sequence length, number of unique tokens, and total number of tokens by corpus, including the cumulative
        total across all corpora.}
\end{table}

@@

Sci-Kit Learn's implementation of the LOF
algorithm was used with a value of \(N = 35\) for \(N\) neighbors in the K-Nearest Neighbors algorithm and Minkowski distance for distance
metrics/calculations (in this case, since \(p=2\), Euclidean distance is effectively used for distance calculations). A total of 47 samples were
detected as outliers and removed for analysis and are summarized in table~\ref{table:outlier_stats}.

\begin{table}[h!]
    \centering
    \begin{tabular}{l | r}
        \toprule
        \multicolumn{2}{c}{Outlier Properties}   \\
        \midrule
        \textbf{Mean Sequence Length}    & 47.30 \\
        \textbf{Std of Sequence Length}  & 11.82 \\
        \textbf{Minimum Sequence Length} & 38    \\
        \textbf{Maximum Sequence Length} & 73    \\
        \textbf{\# Samples}              & 47    \\
        \bottomrule
    \end{tabular}
    \label{table:outlier_stats}
    \caption{Mean, standard deviation, minimum, and maximum sequence length of samples detected as outliers.}
\end{table}

@@

\section{Language Modeling}
% bert, roberta
% ngram
This section introduces the chosen language model architectures as well as their pretrained checkpoints, if one is used. Two transformer-based
architectures are used, BERT and RoBERTa, and statistical/machine learning-based approaches such as N-gram and word lattices.

The transformers are pre-trained using the same masked language modeling (MLM) objective presented in Devlin et al.~\cite{devlin_bert_2019}
to train BERT for bidirectional encoding representation. Both pretrained checkpoints and base models are pretrained and used with the best performing
ASR model to analyze their impacts on performance. The pre-training data is taken from the unified transcripts as described above, tokens are masked
at random with a probability of 0.2. N-grams are trained using the Natural Language Toolkit (NLTK) library in Python with values for $N$ of 1, 2,
and 3.

\section{Speech Recognition}

\section{Tokenization Algorithms}
% present and explain the tokenization algorithms used (possibly which models they correspond to)
Tokenization is the process of separating or segmenting samples into tokens. The rules by which the tokens are created or segmented depends entirely
on the algorithm used. The simplest and most intuitive way to do this by creating tokens based on whitespace i.e., each ``word'' in the corpus
is treated as a token\footnote{The Huggingface Tokenizer library implements this as the \lstinline|WordLevel| tokenizer.}. Although this is a simple
and easy to understand method of tokenization, it does little to nothing to address the out-of-vocabulary problem that necessarily comes with any
NLP tasks~\cite{wu_googles_2016}.

\subsection{WordPiece}

\subsection{(Byte-Level) Byte-Pair Encoding}

>>>>

\usepackage{cite}
\usepackage{listings}
\usepackage{multicol}
\usepackage{newtxtext}
\usepackage{setspace}
\usepackage{sectsty}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{xcolor}
\usepackage[T1]{fontenc}
\usepackage[numbib]{tocbibind}
\usepackage[lmargin=1.25in,rmargin=1.25in,tmargin=1.0in,bmargin=1.0in]{geometry}

\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}

\lstdefinestyle{thesiscodelisting}{
    keywordstyle=\color{magenta},
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{green},
    basicstyle=\ttfamily\footnotesize,
    breakatwhitespace=false,
    breaklines=true,
    captionpos=b,
    keepspaces=true,
    showspaces=false,
    showstringspaces=false,
    showtabs=false,
}

\lstset{style=thesiscodelisting}

@@

    \normalsize
    by\\
    \textit{Aaron Van De Brook}\\

@@

% Signature lines/committee members
\begin{center}
    \begin{minipage}{3in}
        \vspace*{0.25in}
        THESIS COMMITTEE:\\
        \vspace*{0.25in}
        \hrule
        \vspace*{2pt}
        Jianhua Liu, Ph.D.\\
        Committee Chairman\\
        \vspace*{0.75in}
        \hrule
        \vspace*{2pt}
        Prashant Shekhar, Ph.D.\\
        Committee Member\\
        \vspace*{0.75in}
        \hrule
        \vspace*{2pt}
        Andrew Schneider\\
        Committee Member\\
        \vspace*{0.75in}
        \hrule
        \vspace*{2pt}
        Massood Towhidnejad, Ph.D.\\
        Department Chair, Electrical Engineering \& Computer Science\\
        \vspace*{0.75in}
        \hrule
        \vspace*{2pt}
        Associate Vice President for Academics
    \end{minipage}
\end{center}
\newpage
\tableofcontents
\newpage
\doublespacing{}

\section{Notation and Terminology}
% TODO: proper writing, bulleted list for now
This section is included for the sake of clarity throughout this thesis. In pursuit of that, there are a few terms and notations that are defined
below.

Set notation is used in parts of this thesis, particularly in equations and algorithms, to convey mathematical concepts. With that being said, the
shorthand below is defined
\begin{equation*}
    A; x := A \cup \{x\}
\end{equation*}

\noindent
as adding the element $x$, of its own set, to the set $A$.

The following words (in bold) are used throughout the thesis as they relate to language modeling and are defined concretely as follows:
\begin{itemize}
    \item A \textbf{sequence} shall be defined as a set of tokens (distinct from a sentence)
    \item A \textbf{token} shall be defined as the smallest, fundamental component of a sequence (distinct from characters and words)
    \item The rules for the formation of tokens from a sentence (and into sequences) shall be defined by tokenization algorithms (section
          \ref{sec:tokenizers})
    \item The definition of \textbf{words} and \textbf{sentences} follow from that of the English language
    \item The \textbf{vocabulary}, \textbf{inventory}, \textbf{index}, and \textbf{word list} of a model shall be semantically identical
\end{itemize}
\newpage

\section*{Abstract}\label{sec:abstract}

\section{Introduction}\label{sec:introduction}
There have been noticeable and significant advancements in the general domain of artificial intelligence (AI) recently. Automatic speech recognition and
language modeling technologies, in particular, have made tremendous advances over the past two decades achieving performance levels good enough to
be brought to consumer facing products such as personal assistants, search engines, phones, televisions, etc. Due to the success in the general
domain, applications have begun to spin-off into problem-specific domains, such as (and most notably for this work) aviation. For example, several
speech recognition systems with rule-based natural language inference and semantic extraction algorithms have been implemented at EUROCONTROL
simulation centers in Europe to train air traffic controllers and study the potential effects of AI augmented air traffic management (ATM) systems
on controller workloads. Preliminary studies have found that AI augmented ATM systems not only reduce controller workloads but also reduce fuel
consumption of aircraft by optimizing aircraft routes in the airspace and reducing departure/arrival times thereby reducing fuel
consumption~\cite{helmke_quantifying_2017}.

At the time of this writing, the current focus of general machine learning algorithms in the automatic speech recognition (ASR) and language modeling
domains has been on a variety of deep learning methods/architectures, namely; recurrent, convolutional, and transformer neural networks due to the
availability of large corpora such as LibriSpeech~\cite{panayotov_librispeech_2015} and Mozilla Common Voice~\cite{ardila_common_2020} for ASR and
WikiText~\cite{merity_pointer_2016} and IMDB~\cite{maas_learning_2011} corpora for language modeling. In contrast to the general domain, the aviation
domain still prefers, and sees much success with, traditional machine learning models for both ASR and language modeling applications. Typically, the
most prevalent of the traditional models are Hidden Markov models and Gaussian Mixture models, in conjunction with word lattices, N-grams, and/or
rule-based approaches depending on the application and operating environment of the models. The most likely reason for this divergence between the
domains is the difference and relative lack of labeled data in the aviation domain. By comparison, one of the largest corpora in the general domain is
LibriSpeech with approximately 1000 hours of labeled data whereas, for aviation, the largest corpus is Air Traffic Control Complete with approximately
26 hours of labeled data\footnote{Air Traffic Control Complete has approximately 70 hours of audio data, however, about 26 hours of the 70 are
    labeled.}. This disparity in data availability is exacerbated by the fact that aviation English uses a specialized vocabulary in addition to
specialized pronunciation for some words to enhance understandability in a radiotelephonic (R/T) medium~\cite{paltridge_handbook_2013}, making it
significantly more difficult to transcribe/label data as it requires transcribers to have domain knowledge and experience in order to create effective
transcriptions. The medium of communication along with the environment in which the communication occurs i.e.~a noisy and possibly low fidelity
medium\footnote{Depending on the equipment used to transmit and receive R/T communications.} combined with a noisy and active communication
environment---an active flight deck (influenced by wind, engine noise, etc.~) or an active Air Traffic Control (ATC) tower (other people/controllers
conversing or communicating in the background)---and a rapid rate of speech~\cite{paltridge_handbook_2013}, further contributing to the difficulty in
understanding and therefore transcribing aviation English.

% connect complexity of aviation english to utility/necessity of NLP models (ASR and LM) in the aviation domain
The potential for speech recognition and language modeling algorithms to reach better than human performance (for example, the SQuAD benchmark has
recorded results consistently better than human since 2019\footnote{https://rajpurkar.github.io/SQuAD-explorer/})~\cite{zhang_ai_2022} suggests that
these NLP algorithms could have high utility in the aviation domain. If they can achieve better-than-human performance in the aviation domain, and be
used in conjunction with pilot and air traffic controller systems, they could easily boost efficiency, reduce workloads, and reduce and/or mitigate
errors during normal operations.

The methods and results laid out in this work were done with the intention of creating a streamlined process for training and testing deep learning
based NLP methods, specifically automatic speech recognition and text-based language models. The results from the language models suggest that the
volume of text used to train the language models is not yet sufficient to produce results akin to those in the conversational and/or literary English
domains. However, the automatic speech recognition models were able to achieve near state-of-the-art results and, according to the literature review,
better than other deep learning-based approaches in the aviation domain.

@@

\section{Data Sources}\label{sec:data_source}
There are four corpora used in this work. The datasets are primarily speech corpora intended either for automatic speech recognition research in the
domain of aviation communications, linguistics research in aviation English, or both. All data in all four corpora include both recordings and
transcriptions, audio and text, respectively. The transcriptions are used to pretrain and train the language models and tokenizers (described in
sections \ref{sec:language_modeling} and \ref{sec:tokenizers}).

The four corpora that were selected and used for this work are briefly described below. Further analysis is performed in section \ref{sec:data_analysis}.

\subsection{Air Traffic Control Complete}\label{sec:atcc}
The Air Traffic Control Complete (ATCC) corpus is a speech corpus consisting of audio recordings with corresponding transcriptions.
ATCC is a collection of three smaller corpora recorded at Dallas-Fort Worth, Logan International, and Washington National airports and
transcribed by current or former air traffic controllers familiar with the respective areas. Audio data was recorded through the placement of VHF
antennae configured to monitor several frequencies at each airport, such as arrival, departure, approach, ground, etc. The types of frequencies
monitored vary by airport, but the presence of departure, approach, and ground frequencies is relatively consistent across subcorpora~\cite{godfrey_air_1994}.
% TODO: data length/size

\subsection{ATCO2}\label{sec:atco2}
The ATCO2 dataset is a speech corpus that consists of audio communications at Prague and Brno airports, in Czechia, along with corresponding
transcriptions. The speech recordings and transcripts are crowd sourced from volunteers. The dataset was created and labeled, in part, as an English
language detection corpus, however, the labeling also includes transcripts of speech segments which makes it practical for ASR and language model
development in addition to language detection (although in this work it is only used for ASR and LM development)~\cite{szoke_detecting_2021}.
% TODO: data length/size

\subsection{Air Traffic Control Simulation}\label{sec:atcosim}
The Air Traffic Control Simulation (ATCOSIM) corpus is a speech corpus consisting of audio recordings and corresponding transcriptions.
The data was recorded at a air traffic control simulation center in Bretigny-sur-Orge, France, specifically the EUROCONTROL Experimental Centre.
In this data, only the controllers' voices are included and therefore the transcripts only include the controller's side of each interaction.
The audio data was transcribed by one person, trained according to the guidelines and formatting requirements of the corpus. After all data was
transcribed, it was reviewed and corrected and any remaining problems were reviewed by an operational air traffic controller and
resolved\cite{hofbauer_atcosim_2008}.
% TODO: data length/size

\subsection{ZCU CZ ATC Corpus}\label{sec:zcu_atc}
The ZCU CZ ATC corpus consists of audio recordings and corresponding transcripts in the Czech airspace. Both controller and pilot sides of the
communications are included in this data. Annotations were created by experienced transcribers/annotators and labeled samples were randomly selected
for review during the annotation process. After the dataset was completely labeled, all samples were checked, corrected, and
unified\cite{smidl_air_2019}.
% TODO: data length/size

@@

\begin{table}[h!]
    \centering
    \begin{tabular}{l p{0.65\linewidth}}
        \toprule
        Phenomenon               & Definition                                                                                                              \\
        \midrule
        Hesitations              & Instances of hesitation for which the specific realization is not defined (usually some special token is used instead). \\
        \midrule
        Partial \& cut-off words & Partially pronounced words are indicated by a hyphen character where the pronunciation of the word is missing.          \\
        \midrule
        Filled pauses            & Hesitations with a verbal component that has been described or otherwise defined by the corpus/annotator.               \\
        \bottomrule
    \end{tabular}
    \caption{Speech/transcription phenomena with corresponding convention for transcribing or translating, depending on the corpus being processed.}
    \label{tab:phenomena_definitions}
\end{table}

\begin{table}[h!]
    \centering
    \begin{tabular}{l l}
        \toprule
        Phenomenon                   & Convention for representation \\
        \midrule
        Hesitations                  & [HES]                         \\
        \midrule
        Partial \& cut-off words     & -                             \\
        \midrule
        Filled pauses (open-mouth)   & uh, um, ah                    \\
        \midrule
        Filled pauses (closed-mouth) & hm, mhm                       \\
        \bottomrule
    \end{tabular}
    \caption{Conventions for transcript representation/translation of phenomena to keep transcripts consistent across corpora.}
    \label{tab:phenomena_conventions}
\end{table}

@@

\section{Data Analysis}\label{sec:data_analysis}
% sequence lengths, unique tokens, total number of tokens, total number of samples, breakdown by each dataset
Relevant corpora statistics were observed or calculated and are presented in Table~\ref{tab:corpora_stats}. This includes the number of samples in
the corpus, mean sequence length per sample in the corpus, the number of unique tokens in the corpus, and the total number of tokens in the corpus.
The same statistics were also calculated across corpora, since all four corpora will be combined for training and testing the language and speech
recognition models. Note that this is based on the raw data after loading and unifying the corpora, prior to preprocessing for model training/testing
and outlier removal.

\begin{table}[h!]
    \centering
    \begin{tabular}{l r r r r}
        \toprule
        \textbf{Corpus} & \textbf{Samples} & \textbf{Mean Sequence Length (\(\pm \sigma\))} & \textbf{Unique Tokens} & \textbf{Tokens} \\
        \midrule
        ATCC            & 29,862           & \(11.90 \pm 7.22\)                             & 2,209                  & 355,473         \\
        ATCO2           & 874              & \(12.28 \pm 4.12\)                             & 786                    & 10,733          \\
        ATCOSIM         & 9,544            & \(11.23 \pm 4.12\)                             & 823                    & 107,153         \\
        ZCU CZ ATC      & 14,435           & \(10.05 \pm 5.50\)                             & 2,983                  & 145,107         \\
        \midrule
        Total           & 54,715           & \(11.30 \pm 6.36\)                             & 5,040                  & 618,466         \\
        \bottomrule
    \end{tabular}
    \caption{The number of samples, mean sequence length, number of unique tokens, and total number of tokens by corpus, including the cumulative
        total across all corpora.}
    \label{tab:corpora_stats}
\end{table}

@@

Sci-Kit Learn's implementation of the LOF
algorithm was used with a value of \(N = 35\) for \(N\) neighbors in the K-Nearest Neighbors algorithm and Minkowski distance for distance
metrics/calculations (in this case, since \(p=2\), Euclidean distance is effectively used for distance calculations). A total of 47 samples were
detected as outliers and removed for analysis and are summarized in table~\ref{tab:outlier_stats}.

\begin{table}[h!]
    \centering
    \begin{tabular}{l | r}
        \toprule
        \multicolumn{2}{c}{Outlier Properties}   \\
        \midrule
        \textbf{Mean Sequence Length}    & 47.30 \\
        \textbf{Std of Sequence Length}  & 11.82 \\
        \textbf{Minimum Sequence Length} & 38    \\
        \textbf{Maximum Sequence Length} & 73    \\
        \textbf{\# Samples}              & 47    \\
        \bottomrule
    \end{tabular}
    \caption{Mean, standard deviation, minimum, and maximum sequence length of samples detected as outliers.}
    \label{tab:outlier_stats}
\end{table}

@@

\section{Language Modeling}\label{sec:language_modeling}
% bert, roberta
% n-gram
This section introduces the chosen language model architectures as well as their pretrained checkpoints, if one is used. Two transformer-based
architectures are used, BERT and RoBERTa, and statistical/machine learning-based approaches such as N-gram and word lattices.

The transformers are pre-trained using the same masked language modeling (MLM) objective presented in Devlin et al.~\cite{devlin_bert_2019} to train
BERT for bidirectional encoding representation and likewise in Liu et al.~\cite{liu_roberta_2019} for training RoBERTa. Given the results of the
RoBERTa pretraining and downstream task training thereafter, it was initially decided to eliminate the next sentence prediction (NSP) task as it was
used in BERT, however, their results suggest that the success of the MLM-only approach was due to the increase in batch and overall data
size~\cite{liu_roberta_2019}. Taking this into account, along with the very limited amount of data available in the domain of aviation an NSP
pretraining task was implemented where possible\footnote{Only about half of the data (only the data from ATCC, since it is continuous recordings of
    the airspace) is in a format that enables us to extract the turn taking phenomenon that is present in most air-traffic communications.} to improve
its pretraining results. MLM and NSP pretraining are explained more in-depth in section \ref{sec:bert}, below.

\subsection{N-Gram}\label{sec:ngram}
An N-gram language model estimates the probability of a token occurring in a sequence based on the \(N\) tokens preceding or succeeding that token.
In other words, for a sequence, $X$, of $T$ tokens

\begin{equation}\label{eq:ngram_sequence}
    X = \mbox{$w_1$, $w_2$, $w_3$, ..., $w_T$}
\end{equation}

\noindent
the probability of a token, $w_i$, occurring in the sequence can be represented as

\begin{equation}\label{eq:ngram_preceding}
    \begin{gathered}
        P(w_i) = P(w_i|w_{i-1},w_{i-2},w_{i-3},...,w_{i-N}) \\
        P(w_i) = \prod_{j=i-1}^{i-N} P(w_j)
    \end{gathered}
\end{equation}

\noindent
or for the case of succeeding tokens (looking ahead $N$ tokens)

\begin{equation}\label{eq:ngram_succeeding}
    \begin{gathered}
        P(w_i) = P(w_i|w_{i+1},w_{i+2},w_{i+3},...,w_{i+N}) \\
        P(w_i) = \prod_{j=i+1}^{i+N} P(w_j)
    \end{gathered}
\end{equation}

\noindent
where $N$ is the number of preceding (or succeeding) tokens. Since the $N$ preceding tokens\footnote{The most common use case for an N-gram model is
    the preceding tokens.} are needed prior to the estimation of the probability of $w_i$, N-grams are sometimes referred to as left-associative
models (in a left-to-right language).

The probability of a token occurring anywhere in a sequence (referred to as $w_j$ in equations \ref{eq:ngram_preceding} and \ref{eq:ngram_succeeding})
is determined statistically as the number occurrences of a token in a corpus divided by the total number of tokens in the corpus.

\begin{equation}\label{eq:ngram_prob_wj}
    P(w) = \frac{\textit{occurrences}(w)}{D_{tokens}}
\end{equation}

\noindent
where $P(w)$ is the probability of a token, $w$, occurring anywhere in the corpus, $\textit{occurrences}(w)$ is the number of times of a specific
token appears in the corpus, and $D_{tokens}$ is total number of tokens in the corpus, $D$. Computationally, the process of determining the
probabilities of the individual tokens in the corpus is typically referred to as training or fitting the N-gram model.

$N$ is a predetermined value based on model or application requirements and while it can technically take any value from one to infinity, its
usability becomes very limited at the minimum sequence length in the corpus and unusable at any value at or above the maximum sequence length.
Depending on implementation, the time and resources required to train a model with a large value of $N$ can be very expensive. The most common values
for $N$ in the literature are one, two, and three, sometimes called unigrams, bigrams, and trigrams, respectively.

\subsection{BERT}\label{sec:bert}
The purpose of BERT (Bidirectional Encoder Representations from Transformers) is to represent tokens in text sequences in relation to the tokens
preceding and following the token in question. Whereas an N-gram model represents token probabilities based on the $N$ tokens preceding a specific
token or, conversely, the $N$ tokens succeeding said token, BERT models are able to predict the probability of a token based on both preceding and
succeeding tokens, thus the bidirectionality novelty.

\textbf{Masked Language Modeling}. The ability of BERT to represent text in a bidirectional context is achieved through the Masked Language Modeling
(MLM) pretraining task~\cite{devlin_bert_2019}. MLM pretraining choses a certain percentage of token positions at random for prediction, then tokens
are replaced with either a ``mask'' token, a random token, or the original token according to preconfigured probabilities (BERT used 80\% for the mask
token, 10\% for random tokens, and 10\% for the original/unchanged token). The original token is then predicted with cross-entropy loss.

\textbf{Next Sentence Prediction}. NSP is meant to support downstream tasks, while MLM conditions the model to represent tokens within sentences, NSP
is meant to condition the model to understand relationships between sentences/sequences. Since BERTs training set consists primarily of literary works
and excerpts from Wikipedia, the model simply predicts whether two sentences appear together (one after the other) in the original text. Instead of a
single input sequence, the input consists of two sequences separated by a ``separation'' token and beginning with a ``classification'' token, these
are represented by \lstinline|[SEP]| and \lstinline|[CLS]|, respectively in Devlin et al.~\cite{devlin_bert_2019}.

For symmetry between pretraining tasks, the special tokens specifying the beginning and end of input sequences are shared between tasks. For both MLM
and NSP objectives, input sequences always start with a classification token and end with a separation token (see figure \ref{fig:bert_mlm_input_example}).

\begin{figure}[h!]
    \includegraphics[width=\linewidth]{figures/BERT_MLM_input.png}
    \caption{An example of an input sequence for a masked language modeling pretraining task. Modified from Devlin et al.~\cite{devlin_bert_2019}.}
    \label{fig:bert_mlm_input_example}
\end{figure}

\noindent
Likewise, NSP inputs always begin with a classification token and each input sequence is concatenated with a separation token terminating each
sequence (see figure \ref{fig:bert_nsp_input_example}).

\begin{figure}[h!]
    \includegraphics[width=\linewidth]{figures/BERT_NSP_input.png}
    \caption{An example of an input sequence for a next sentence prediction pretraining task.}
    \label{fig:bert_nsp_input_example}
\end{figure}

\subsection{RoBERTa}\label{sec:roberta}
As the name implies, RoBERTa, is an iteration on BERT (described in section \ref{sec:bert}, above). The results from
Liu et al.~\cite{liu_roberta_2019} indicate that BERT was initially significantly undertrained, so by training longer, with larger batch sizes, on
longer sequences, and dynamically changing the masking pattern applied to the MLM pretraining task, the performance of BERT was improved. The new
checkpoint of this version of BERT was therefore called RoBERTa (Robustly Optimized BERT Pretraining Approach).

The pretraining methodology introduced by RoBERTa eliminated the NSP pretraining task that was originally used in BERT and opted instead to increase
the quantity of training data along with the batch sizes in addition to training the model longer. Lastly, there is a small difference in the
construction of input sequences as compared to BERT; as with BERT, input sequences always begin with a classification token (\lstinline|[CLS]|) and
individual segments are separated by a separation token (\lstinline|[SEP]|). Rather than having sequences end with a separation token as well, they
are terminated with an end of sequence token (\lstinline|[EOS]|). See equation \ref{eq:roberta_input_example} modified from Liu et
al.~\cite{liu_roberta_2019}, below.

\begin{equation}\label{eq:roberta_input_example}
    \mbox{[CLS], $x_1$, $x_2$, ..., $x_N$, [SEP], $y_1$, $y_2$, ..., $y_M$, [EOS]}
\end{equation}

% \section{Speech Recognition}\label{sec:speech_recognition}
% Automatic speech recognition (ASR) or simply speech recognition is the process of translating spoken language to text i.e.~transcribing spoken
% language. There have been a myriad of approaches to ASR problems, including, but not limited to rule-based, machine learning, and deep learning
% methods. For the purposes of this work, we will focus primarily on deep learning approaches to ASR tasks with a specific focus on transcribing
% spoken language in aviation (following from the data accumulated, processed, and analyzed in sections \ref{sec:data_source},
% \ref{sec:data_processing}, and \ref{sec:data_analysis}).

% Three ASR architectures were chosen due to their success in the conversational English domain and ease of reproducing results. Nvidia maintains a
% Python package called Neural Modules (shortened to NeMo) with PyTorch implementations of many well-performing ASR models\footnote{Performance of these
%     models can be seen in their respective papers (referenced hereafter).}, including the three models used in this work and described in more detail
% below \cite{kuchaiev_nemo_2019}.

% Although the internal processing of data varies between architectures, the format of the models input and output is generally the same. ASR models
% have seen much success and performance gains through the use of Mel-frequency Cepstral Coefficients (MFCCs), so the audio files provided by the
% corpora used in this work were processed and resampled into the same format, then the MFCCs are extracted according to the process
% in~\cite{sahidullah_design_2012}. The MFCCs are then used as input to the ASR models. Each model outputs a probability distribution of characters or
% tokens (depending on the vocabulary the ASR model is initialized with) over a segment of audio. All the ASR models used are trained with a
% Connectionist Temporal Classification (CTC) loss function~\cite{graves_connectionist_2006}.

% \subsection{Jasper}\label{sec:jasper}
% Jasper is a family of end-to-end convolutional neural networks (CNNs) developed by NVIDIA to replace traditional ASR models that use separately
% learned components for each stage of the pipeline (acoustic, pronunciation, language modeling, etc.) \cite{li_jasper_2019}.

% Jasper models are designed with $B$ blocks and $R$ sub-blocks and are named accordingly as \textit{Jasper $B\mbox{x}R$} models. The Jasper
% architecture also contains four extra convolutional blocks, one for preprocessing and three for postprocessing (see figure \ref{fig:jasper}). Each
% convolutional sub-block contains one dimensional convolutions. Li et al.~trained and tested a Jasper 10x5 model that produced state-of-the-art
% results on the LibriSpeech, Wall Street Journal, and Fisher+Switchboard speech corpora \cite{li_jasper_2019}.

% \begin{figure}[h!]
%     \includegraphics[width=\linewidth]{figures/jasper_vertical.png}
%     \caption{Block diagram of the Jasper architecture. Copied from Li et al.~\cite{li_jasper_2019}.}
%     \label{fig:jasper}
% \end{figure}

% \subsection{QuartzNet}\label{sec:quartznet}
% The QuartzNet architecture is based on Jasper with some key modifications, namely, that the one dimensional convolutions are replaced with ``one
% dimensional time-channel separable convolutions''. As with Jasper, QuartzNet models are denoted as \textit{QuartzNet $B\mbox{x}R$}, however, blocks
% are repeated $S$ times and have a certain number of input and output channels, $c_{in}$ and $c_{out}$, respectively.
